---
title: "Barn Owl Gene Expression – BIOL 710"
author: "Pinkal Tandel"
output:
  pdf_document: default
  html_document: default
---
## Introduction

Malaria is the most common parasitic disease in humans, transmitted by the bite of anopheline mosquitoes carrying Plasmodium spp. parasites [1]. It affects hundreds of millions globally, with an estimated 263 million cases in 2023, resulting in 597,000 deaths across 83 countries [2]. Plasmodium parasites infect red blood cells and belong to the order Haemosporidia, which also includes avian haemosporidians such as Leucocytozoon species. These parasites cause malaria-like symptoms in birds and have been associated with the decline of certain bird species [3]. Leucocytozoon infections are known to induce significant immune responses and physiological stress in birds.

While extensive research has been conducted on haemosporidian parasites, the role of viruses within these protozoa has remained understudied. Avian haemosporidian species show variability in their effects on hosts, with some infections being nearly undetectable, while others cause severe anemia and death. Avian malaria caused by Leucocytozoon parasites is a widespread disease in birds, often leading to significant physiological stress and immune responses in infected hosts.

In 2019, a new class of RNA viruses infecting parasitic protozoans was identified, characterized by a bi-segmented genome consisting of an RNA-dependent RNA polymerase (RdRp) and a hypothetical protein with two overlapping open reading frames (ORFs) of unknown function. The concept of “Matryoshka” infections, where parasites themselves contain viral infections, has been increasingly studied in protozoan biology [4]. These viruses, named Matryoshka RNA viruses (MaRNAV), were first discovered in Plasmodium vivax-infected humans and mosquitoes in eastern Malaysia, with the initial strain designated as MaRNAV-1. Subsequent studies using meta-transcriptomic analysis identified MaRNAV-2 in Leucocytozoon-infected Australian birds [5,6]. These findings suggest that viruses may influence the biology of avian Leucocytozoon parasites, raising important questions about their impact on host-pathogen interactions and disease outcomes.

Further research revealed additional MaRNAV strains in birds infected with haemosporidians: MaRNAV-3 and MaRNAV-6 were detected in American birds carrying Leucocytozoon species, while MaRNAV-4 and MaRNAV-5 were associated with Haemoproteus infections [6]. These findings highlight a broader parasite-virus relationship within avian malaria systems and emphasize the need to investigate host immune responses in co-infection contexts.

Co-infection is increasingly recognized as a key factor in shaping protozoan infections, with evidence suggesting that viral infections in parasites can modulate virulence, immune evasion, and transmission dynamics. Understanding host-pathogen interactions could provide insight into avian immune responses, and understanding how immune genes especially innate immune genes that serve as the initial defense mechanism have evolved in birds, a highly diverse group of tetrapods, is crucial [7].

Co-infections like these may influence parasite behavior and host immune response, but the biological significance of MaRNAVs is still unknown. In avian malaria systems, understanding these interactions could provide valuable insights into immune evolution and disease outcomes.

Barn owls (Tyto alba) were chosen for this study as a model species, as they have been previously utilized in research to assess immune responses in birds infected with Leucocytozoon and MaRNAV. This choice allows for control over species-specific factors, providing a standardized approach to studying gene expression in response to haemosporidian infection. This project builds upon a prior RNA-seq study that identified immune-related genes differentially expressed in infected owls. That study, however, was limited to only six individuals. In an earlier study by Caroline Faircloth, RNA-seq analysis of barn owls (Tyto alba) identified genes that were differentially expressed in owls infected with both Leucocytozoon and MaRNAV.To build on this foundational work, this project focuses on expanding the dataset by adding more Leucocytozoon-infected barn owls and analyzing expression of a small number of key immune genes for validation.

To strengthen those findings, this study adds new Leucocytozoon-infected barn owl replicates to the dataset and focuses on validating a small number of immune-related genes such as, MAP4K4, HEBP1 previously identified. The ultimate goal is to prepare these targets for downstream ddPCR validation and support the completion of a publishable manuscript.


## Research Question

Which immune-related genes are upregulated in barn owls infected with Leucocytozoon, and can we strengthen RNA-seq results through replication and ddPCR validation?

### Aims

Aim 1: Expand the Dataset
Increase the number of barn owl samples infected with Leucocytozoon, building on the original RNA-seq study. This improves statistical power and robustness of observed gene expression differences.

Aim 2: Focus on Key Immune Genes
Analyze expression levels of a small number of immune-related genes such as HEBP1, MAP4K4, TLR7, IFNG identified in prior work. Prepare for downstream ddPCR validation of these genes in additional owl samples.

## Sample Metadata

| Sample ID | Condition  | Age      | Notes                            |
|-----------|------------|----------|----------------------------------|
| LWC142    | Positive   | Juvenile | Minor foot wounds; eye irritation |
| LWC143    | Positive   | Adult    | Minor body wounds; eye irritation |
| LWC153    | Positive   | Juvenile | Fell from nest                   |
| LWC148    | Negative   | Juvenile | Euthanized (fracture)            |
| LWC157    | Negative   | Juvenile | Found in dumpster                |
| LWC162    | Negative   | Juvenile | Found on ground                  |

- “Positive” = *Leucocytozoon*-infected
- “Negative” = uninfected with both parasite and virus (per Caroline’s RT-PCR and RNA-seq screening)

## Repository Structure
- `data/`: Contains gene expression CSV and DatasetCard.md
- `owl_gene_analysis.Rmd`: Main R Markdown file with analysis and interpretation
- `README.md`: Project overview

## Notes

This work supports a follow-up ddPCR experiment to quantify expression levels of key immune genes. Genes were selected based on differential expression in Caroline’s paper and relevance to *Leucocytozoon* or *Plasmodium* infection studies in birds.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load merged gene expression data from 6 barn owls
gene_data <- read.csv("data/gene_expression_data.csv")

# Preview the first few rows
head(gene_data)

```

```{r}
# Number of genes and samples
nrow(gene_data)   # Total number of genes
ncol(gene_data)   # Number of samples + GeneID column

# Check for missing values
sum(is.na(gene_data))   # Should be 0

```

```{r}
# View expression counts for MAP4K4
gene_data[gene_data$GeneID == "MAP4K4", ]

```

## References

1. Ferraboli JW, Veiga GTS da, Albrecht L. Plasmodium vivax transcriptomics: What is new? Exp Biol Med. 2023;248(19):1645. doi:10.1177/15353702231198070

2. Fact sheet about malaria. Accessed March 7, 2025. https://www.who.int/news-room/fact-sheets/detail/malaria

3. Valkiūnas G. Avian Malaria Parasites and Other Haemosporidia. CRC Press; 2005.

4. Rodrigues JR, Roy SW, Sehgal RNM. Novel RNA viruses associated with avian haemosporidian parasites. PloS One. 2022;17(6):e0269881. doi:10.1371/journal.pone.0269881

5. Petrone ME, Charon J, Grigg MJ, et al. A virus associated with the zoonotic pathogen Plasmodium knowlesi causing human malaria is a member of a diverse and unclassified viral taxon. Published online September 19, 2024. doi:10.1101/2024.09.18.613759

6. Duc M, Esperanza C, Chagas CRF, Iezhova T, Sehgal RNM, Valkiūnas G. First report of Matryoshka RNA virus in an African-European migrant bird. PLOS ONE. 2025;20(3):e0319395. doi:10.1371/journal.pone.0319395

7. Outlaw DC, Walstrom VW, Bodden HN, Hsu C Yu, Arick M, Peterson DG. Molecular evolution in immune genes across the avian tree of life. Parasitol Open. 2019;5:e3363. doi:10.1017/pao.2019.33
